---
title: Baseline and In-Hospital Risk Factors for Mortality and Other Outcomes among
  TBI Patients
author: 'Jennifer Thompson, MPH; Supervisor: Rameela Chandrasekhar, PhD'
date: '`r format(Sys.time(), "%B %d %Y")`'
output:
  html_notebook:
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float: yes
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: 2
abstract: The following analyses examine baseline and in-hospital risk factors for
  in-hospital mortality, in-hospital and discharge characteristics, and long-term
  mortality among a cohort of traumatic brain injury (TBI) patients with at least
  one day in the ICU.
---

```{r knitrsetup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(width = 200)
```

```{r setup, message = FALSE, results='hide'}
## Make indentation for Markdown tables easier
space.3 <- paste(rep("&nbsp;", 3), collapse = '')
space.6 <- paste(rep("&nbsp;", 6), collapse = '')
space.9 <- paste(rep("&nbsp;", 9), collapse = '')
space.12 <- paste(rep("&nbsp;", 12), collapse = '')
space.15 <- paste(rep("&nbsp;", 15), collapse = '')

## Load libraries and original analysis data sets
library(captioner)
library(tidyverse)
library(pander)
library(rms)

## -- Set up figure, table numbering using the captioner package -----------------------------------
table_nums <- captioner(prefix = 'Table')
table_nums(name = 'consort', caption = 'Cohort Inclusion and Exclusion')
table_nums(name = 'descstats_oneobs',
           caption = 'Baseline, In-Hospital, & Discharge Characteristics')
table_nums(name = 'descstats_daily', caption = 'Daily In-Hospital Characteristics')
table_nums(name = 'clinical_imputation',
           caption = 'Missingness among Daily Variables Imputed by Rule')
table_nums(name = 'mortality_patients', caption = 'Inclusion in Mortality Models')
table_nums(name = 'mortality_missing', caption = 'Missing Data by Patient-Day for Mortality Models')

table_nums(name = 'package_info', caption = 'R Packages Loaded')

## Load analysis data sets
load('AnalysisData/tbi_datasets.Rdata')

## -- Create analysis data sets including only ICU patients ----------------------------------------
## Patients with an ICU LOS of at least 1 day, admitted with blunt trauma
icu.pts <- unique(filter(tbi.oneobs, !is.na(icu.los) & icu.los > 0 & pt.injury == 'Blunt')$mrn)

## Filter daily data
tbi.daily.icu <- filter(tbi.daily, mrn %in% icu.pts)

## How many patients who would otherwise be included had no daily EMR data?
icu.nodaily.pts <- setdiff(icu.pts, unique(tbi.daily.icu$mrn))

## MRNs of all patients included in final analyses (>=1 ICU day, demographic + daily data)
pts.inc <- intersect(icu.pts, unique(tbi.daily.icu$mrn))

## Subset tbi.oneobs to only include pts.inc
tbi.oneobs.icu <- filter(tbi.oneobs, mrn %in% pts.inc)

```

# Cohort Description
We originally collected electronic medical record data for `r length(unique(tbi.oneobs$mrn))`
patients aged 16 years or older admitted to VUMC between August 2006 and July 2012 with traumatic
brain injury (TBI) requiring hospitalization. TBI was classified based on ICD-9 diagnostic codes,
and includes concussion without intracranial hemorrhage. Code used to create analysis data sets from
raw EMR data can be found
[here](https://github.com/jenniferthompson/TBIMort/blob/master/tbimortality_datamgmt.R).

For these analyses, we focused on patients admitted to the ICU and who had blunt (vs penetrating)
trauma. `r table_nums('consort', display = 'cite')` describes how patients were included and
excluded from our broad cohort, with requirements for inclusion in **bold**. Patients included in
each specific analysis will be described in each section, beginning with this group of
`r length(pts.inc)`.

```{r createconsorttable}
## -- Calculate Ns for each group of patients included/excluded ------------------------------------
## Original number of TBI patients
n.alltbi <- length(unique(tbi.oneobs$mrn))

## Had penetrating vs blunt trauma
n.injuryna <- sum(tbi.oneobs$pt.injury == 'Missing', na.rm = TRUE)
n.penetrating <- sum(tbi.oneobs$pt.injury == 'Penetrating', na.rm = TRUE)
n.blunt <- sum(tbi.oneobs$pt.injury == 'Blunt', na.rm = TRUE)

## Blunt injury and had ICU LOS >= 1 day?
n.nolos <- sum(is.na(subset(tbi.oneobs, pt.injury == 'Blunt')$icu.los))
n.noicu <- with(subset(tbi.oneobs, pt.injury == 'Blunt'), sum(!is.na(icu.los) & icu.los == 0))
n.icu <- length(icu.pts)

## No daily data vs included in main analyses (meaning, at least one day of daily data)
n.nodaily <- length(icu.nodaily.pts)
n.included.main <- length(pts.inc)

## Create CONSORT table
consort.table <- data.frame(Patients = c('All TBI Patients',
                                         paste0(space.3, 'Missing injury type'),
                                         paste0(space.3, 'Penetrating injury'),
                                         paste0(space.3, '**Blunt injury**'),
                                         paste0(space.6, 'Missing ICU LOS data'),
                                         paste0(space.6, '0 days in ICU'),
                                         paste0(space.6, '**>=1 day in ICU**'),
                                         paste0(space.9, 'Missing daily EMR data'),
                                         paste0(space.9, '**Eligible for inclusion**')),
                            N = c(n.alltbi, n.injuryna, n.penetrating, paste0('**', n.blunt, '**'),
                                  n.nolos, n.noicu, paste0('**', n.icu, '**'), n.nodaily,
                                  paste0('**', n.included.main, '**')))

```

#### `r table_nums('consort')`
```{r printconsorttable, results = 'asis'}
pandoc.table(consort.table, justify = c('left', 'right'))

```
#### `r table_nums('descstats_oneobs')`
```{r printdescstatsoneobs, results = 'markup'}
summaryM(age + gender + race + insurance.code + iss + cpr.yn + pt.marshall + pt.cerebral +
           pt.epidural + pt.injury + vent.days + days.mental + days.del + days.coma + dcfd.14 +
           icu.los + los + hosp.death + disposition.coded + fim.total + died.3yr ~ 1,
         data = tbi.oneobs.icu)

```

#### `r table_nums('descstats_daily')`
```{r printdescstatsdaily, results = 'markup'}
summaryM(mental.status + max.motor.imp + pupil.react.imp + min.glucose.imp + min.hemoglobin.imp +
           min.sodium.imp + max.icp.imp + max.icp.dich + sofa.nanormal.imp + tot.benzo +
           tot.opioid + tot.propofol + tot.dex + tot.antipsyc + tot.betablock + tot.pento +
           tot.clonid + units.cryo + units.plasma + units.platelets + units.prbc ~ 1,
         data = tbi.daily.icu)

```

# Clinical Imputation Rules for Missing Data
As this is medical record data, there is a large amount of missingness: labs are not ordered unless
there is reason to order them, for example. We developed a set of clinical imputation rules to deal
with much of this missingness among our covariates.

Several daily variables were imputed by using the closest value within two days before or after the
missing value. If two values equidistant from the missing day were available, preference was given
to the earlier value. `r table_nums("clinical_imputation", display = 'cite')` shows the number of records originally missing for each of these variables; the number successfully imputed using this two-day rule; and the number which remain missing after the two-day rule was implemented.

```{r clinicalimp}
## -- For each covariate imputed via two-day rule, get N missing in original data, -----------------
## -- N successfully imputed, and N still missing after imputation ---------------------------------
clinical.imp.covars <- c('max.motor', 'pupil.react', 'min.glucose', 'min.hemoglobin', 'min.sodium',
                         'sofa.resp', 'sofa.cns', 'sofa.cv', 'sofa.coag', 'sofa.renal')

clinical.imp.npct.missing <- do.call(rbind,
                                     lapply(clinical.imp.covars, FUN = function(vname){
  vname.imp <- paste0(vname, '.imp')
  
  ## N, % missing in original data
  nmiss.org <- sum(is.na(tbi.daily.icu[,vname]))
  pctmiss.org <- round((nmiss.org / nrow(tbi.daily.icu))*100)
  
  ## N, % of missing imputed
  nimp <- sum(is.na(tbi.daily.icu[,vname]) & !is.na(tbi.daily.icu[,vname.imp]))
  pctimp <- round((nimp / nmiss.org)*100)
  
  ## N, % of patient-days still missing after imputation
  nmiss.imp <- sum(is.na(tbi.daily.icu[,vname.imp]))
  pctmiss.imp <- round((nmiss.imp / nrow(tbi.daily.icu))*100)
  
  c("% (N) Missing,\nOriginal Data" =
      paste0(pctmiss.org, '% (', format(nmiss.org, big.mark = ','), ')'),
    "% (N) of Missing\nValues Imputed" = paste0(pctimp, '% (', format(nimp, big.mark = ','), ')'),
    "% (N) Still Missing\nAfter Imputation" =
      paste0(pctmiss.imp, '% (', format(nmiss.imp, big.mark = ','), ')'))
}))

rownames(clinical.imp.npct.missing) <- clinical.imp.covars

## How many records had missing ICP / ICP > 120 in original data?
n.icpmiss <- sum(is.na(tbi.daily.icu$max.icp))
pct.icpmiss <- round((n.icpmiss / nrow(tbi.daily.icu))*100)
npct.icpmiss <- paste0(pct.icpmiss, '% (', format(n.icpmiss, big.mark = ','), ')')

n.icp120 <- sum(tbi.daily.icu$max.icp > 120, na.rm = TRUE)
pct.icp120 <- round((n.icp120 / nrow(tbi.daily.icu))*100, digits = 2)
npct.icp120 <- paste0(pct.icp120, '% (', format(n.icp120, big.mark = ','), ')')

```

#### `r table_nums("clinical_imputation")`
Percentages for original and post-imputation missingness are out of the `r nrow(tbi.daily.icu)` in-hospital patient-days among those patients eligible for inclusion in our analyses.

```{r printclinicalimp, results = 'asis'}
panderOptions('table.emphasize.rownames', FALSE)
panderOptions('keep.line.breaks', TRUE)
pander(clinical.imp.npct.missing, keep.line.breaks = TRUE)

```

Additionally:

- The liver component of the SOFA score was considered to be normal if no bilirubin was measured.
- Missing maximum ICP was assumed to be normal, and assigned a random value from a uniform distribution ranging from 0 to 20 (normal range; imputed values were rounded, to result in integer scores). Maximum ICP > 120 was truncated to 120. In the original data, `r npct.icpmiss` records were missing ICP and assigned a random normal-range value, and `r npct.icp120` had ICP > 120.
- When calculating final full and modified SOFA scores, components which were still missing after
clinical imputation were assumed to be normal (component score = 0). For example, if a patient was hospitalized for five days and never had creatinine measured, that patient would be assumed to have no renal dysfunction during this hospitalization and would have SOFA renal scores of 0 on all five days (though the overall SOFA score would depend on the other components).
- If no dose was recorded for medications or blood products on a given day, we assume the patient received no drug/product, and the dose is considered to be 0.
- If no maximum motor score was available on Days 00 - 02, the *baseline* maximum motor score was considered to be normal and assigned a value of 6.

There is still missing data after these imputation rules are implemented; for example, if a patient's glucose was never measured, there was never a value to carry forward or backward. Specific
amounts of missingness and strategies for handling it will be detailed separately for each analysis.


# Software Details
`r devtools::session_info()$platform$version` was used for all analyses. `r table_nums('package_info', display = 'cite')` shows all add-on packages loaded.

```{r technical, results = 'asis'}
pander(devtools::session_info()$packages)
```
