---
title: "TBI/Mortality Descriptive Statistics"
author: "Jennifer Thompson, MPH; Supervisor: Rameela Chandrasekhar, PhD"
date: "February 10, 2016"
output: pdf_document
header-includes:
   - \usepackage{setspace, relsize}
geometry: margin=0.5in
---

Descriptive statistics are presented below for all variables used in analyses for the TBI mortality
cohort. The first section presents descriptives for baseline and summary characteristics for
each patient; the second presents descriptives for variables collected on a daily basis.

```{r setup, message = FALSE}
library(Hmisc)
load('tbi_datasets.Rdata')
```

All code used to create analysis data sets can be found at [https://github.com/jenniferthompson/TBIMort/blob/master/tbimortality_datamgmt.R](https://github.com/jenniferthompson/TBIMort/blob/master/tbimortality_datamgmt.R). The datasets were created at `r datasets.created.at`,  using raw data exported from REDCap.

# Baseline and Summary Variables

```{r desconeobs, results = 'asis'}
latex(describe(tbi.oneobs), file = '', label = 'table:oneobs')
```

$\pagebreak$

# Daily Variables

```{r descdaily, results = 'asis'}
latex(describe(tbi.daily), file = '', label = 'table:daily')
```
